SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-12-06 papers/B1555_EdgeOn/B1555_letter.tex
Updates to the paper text
Still need to modify the figure

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, as can be seen by comparing the radio contours to the $K^\prime$-band emission  (Fig.~\ref{fig:merlin}), and the faint, but clearly visible emission from {\bf the lensing galaxy what appears to be an edge-on disc}, with a position angle of $\sim 10$~deg and an ellipticity $e \sim 0.8$, where $e$ is defined in terms of the axis ratio $q$ as $q=1-e$.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission that is seen from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in size. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). We immediately see from the high resolution VLBA imaging that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The flux-ratio anomaly and relative sizes of the images at radio-wavelengths implies that there must be some form of perturbation to the gravitational lens mass model, which has thus far been attributed to dark matter substructure within the lensing galaxy or along the line of sight to the lensed images \citep{Dalal2002,Xu12,Xu15}. However, our high-resolution near infrared imaging suggests the intriguing possibility that the mass perturbation could be due to an edge-on disc component within the lens. To test this possibility, we model the system with a disc component in order to see if a plausible mass model can explain the flux-ratio anomaly without the need for additional substructure.

@@

The next step is to test if a physically plausible edge-on disc can cause a perturbation in the strong lensing mass model that produces the observed flux ratio anomaly of  merging images A and B. An exponential disc profile best describes the disc component in spiral galaxies because it matches the light distribution \citep{Kee98}. We thus choose a SIE plus an exponential disc model (SIE+ExpDisc) in our study of B1555+375. The free parameters of the model are the SIE Einstein radius ($b$), centroid position, ellipticity ($e$), position angle ($\theta$), the exponential disc intrinsic central density ($\kappa_0$), and scale length ($R_s$). A detailed definition of these mass profiles is described by \citet{Kee01}. The source position adds two additional model parameters. {\bf We have not included a third component such as an external shear or additional NFW halo term because the number of constraints provided by the compact images detected by MERLIN is not sufficient to constrain such a model. An SIE profile is also sufficient to include the halo mass since B1555 has a small Einstein radius.} Significantly deeper observations at a higher bandwidth with current very long baseline interferometric arrays can test more complex models in the future by imaging the extended gravitational arc and more significantly detecting image D.

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:model}. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total mass within the Einstein radius (1.7~kpc for a lens redshift of $z_l = 1$), which is not unreasonable.  {\bf In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the convergence of the ExpDisc component as a good description of the baryonic mass of lens galaxy.}  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models need the full extent of the gravitational arc for constraints. {\bf The extinction in the AO imaging is so severe that components B and D are not visible, and the arc that connects A and B appears to be truncated.  The low signal-to-noise ratio and the strong extinction in lens galaxy prevent us from using modeling techniques that incorporate the extended NIR emission \citep[e.g.][]{V12}.  Future high resolution radio imaging is needed in order to place stronger constraints on the mass model.}

@@

\textbf{
Here, we find evidence for an edge-on disc in the flux-ratio
  anomaly gravitational lens B1555+375 that can also account for the
  required perturbation in the mass model.  Moreover, near-infrared AO imaging of other lenses in the SHARP sample shows that B1555+375 is not the only system having a significant disc component detection in the lensing galaxy; these gravitational lenses will be investigated in future work, but the images alone suggest that other systems may have similar links between flux-ratio anomalies and disc components. Such high-order deviations
  from a smooth mass distribution have already been suggested
  previously \citep{evans03, congdon05}, but this is the first time
  that an observed disc has been explicitly modelled to account for a
  flux-ratio anomaly. However, it is important to realise that by
  including observationally-motivated additional components/modification[?] to a `smooth' mass
  distribution, such as luminous dwarf companions or edge-on discs, in
  the mass models, we are still limited in our interpretation since we
  must assume that the anomalies are due exclusively to these
  additional components. For B1555+375, it may be possible to independently constrain the mass of the disc through kinematic measurements (most likely from mm-wavelength resolved emission line observations), which 	would test whether the disc makes a sufficient contribution to the mass model \citep[e.g. as in the case of 	B1933+503;][]{suyu12}. Also, in the blue contours of Fig.~\ref{fig:merlin}, B1555+375 shows evidence for an extended $\sim$ 100 mas gravitational arc at radio wavelengths that crosses the edge-on disc. Further high resolution imaging with very long baseline interferometry will provide the observational constraints required to fully test whether the disc is sufficient to cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account 
\citep[e.g.][]{Xu15}. This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.}
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.

>>>>

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, as can be seen by comparing the radio contours to the $K^\prime$-band emission  (Fig.~\ref{fig:merlin}), and the faint, but clearly visible emission from {\bf the lensing galaxy that appears to be an edge-on disc}, with a position angle of $\sim 10$~deg and an {\bf axis ratio $q \sim 0.2$}.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission that is seen from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in {\bf length}. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). We immediately see from the high resolution VLBA imaging that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The flux-ratio anomaly and relative sizes of the images at radio-wavelengths imply that there must be some form of perturbation to the gravitational lens mass model, which has thus far been attributed to dark matter substructure within the lensing galaxy or along the line of sight to the lensed images \citep{Dalal2002,Xu12,Xu15}. However, our high-resolution near infrared imaging suggests the intriguing possibility that the mass perturbation could be due to an edge-on disc component within the lens. To test this possibility, we model the system with a disc component in order to see if a plausible mass model can explain the flux-ratio anomaly without the need for additional substructure.

@@

The next step is to test if a physically plausible edge-on disc can cause a perturbation in the strong lensing mass model that produces the observed flux ratio anomaly of  merging images A and B. An exponential disc profile best describes the disc component in spiral galaxies because it matches the light distribution \citep{Kee98}. We thus choose a SIE plus an exponential disc model (SIE+ExpDisc) in our study of B1555+375. The free parameters of the model are the SIE Einstein radius ($b$), centroid position, ellipticity ($e$), position angle ($\theta$), and the exponential disc intrinsic central density ($\kappa_0$) and scale length ($R_s$). A detailed definition of these mass profiles is described by \citet{Kee01}. The source position adds two additional model parameters. {\bf We have not included a third component such as an external shear or additional NFW halo term because the number of constraints provided by the compact images detected by MERLIN is not sufficient to constrain such a model.} 
%An SIE profile is also sufficient to include the halo mass since B1555 has a small Einstein radius.} 
Significantly deeper observations at a higher bandwidth with current very long baseline interferometric arrays can test more complex models in the future by imaging the extended gravitational arc and more significantly detecting image D.

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:model}. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total {\bf projected} mass within the Einstein radius (1.7~kpc for an assumed lens redshift of $z_l = 1$).  {\bf In this central region of an edge-on spiral galaxy, the baryonic mass can be a significant component of the total enclosed mass \citep[e.g.][]{swells2}.  In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the convergence of the ExpDisc component as a good description of the baryonic mass of the lens galaxy.}  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models need the full extent of the gravitational arc for constraints. {\bf The extinction in the AO imaging is so severe that components B and D are not visible, and the arc that connects A and B appears to be truncated.  The low signal-to-noise ratio and the strong extinction in lens galaxy prevent us from using modeling techniques that incorporate the extended NIR emission \citep[e.g.][]{V12}.  Future high resolution radio imaging is needed in order to place stronger constraints on the mass model.}

@@

\textbf{ Here, we find evidence for an edge-on disc in the flux-ratio
  anomaly gravitational lens B1555+375 that can also account for the
  required perturbation in the mass model.  Moreover, near-infrared AO
  imaging of other lenses in the SHARP sample shows that B1555+375 is
  not the only system having a significant disc component detection in
  the lensing galaxy; these gravitational lenses will be investigated
  in future work, but the images alone suggest that other systems may
  have similar links between flux-ratio anomalies and disc
  components. Such high-order deviations from a smooth mass
  distribution have already been suggested previously \citep{evans03,
    congdon05}, but this is the first time that an observed disc has
  been explicitly modelled to account for a flux-ratio
  anomaly. However, it is important to realise that by including
  observationally-motivated modifications to a
  `smooth' mass distribution, such as luminous dwarf companions or
  edge-on discs, in the mass models, we are still limited in our
  interpretation if we can not determine how much of the anomalies are
  due to these components. For our current analysis of B1555+375, we have
  shown that the entire anomaly could be caused by the edge-on disc.  However,
  it may be
  possible to independently constrain the mass of the disc through
  kinematic measurements (most likely from mm-wavelength resolved
  emission line observations), which would test whether the disc actually makes
  a sufficient contribution to the mass model \citep[e.g. as in the
    case of B1933+503;][]{suyu12}. Also, in the blue contours of
  Fig.~\ref{fig:merlin}, B1555+375 shows evidence for an extended
  $\sim$ 100 mas gravitational arc at radio wavelengths that crosses
  the edge-on disc. Further high resolution imaging with very long
  baseline interferometry will provide the observational constraints
  required to fully test whether the disc is sufficient to cause the
  anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies in some systems. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account 
\citep[e.g.][]{Xu15}. The SWELLS survey, for example, presents a sample of 16 edge-on disks from SDSS \citep{swells1}.  This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.}
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.
